Rclean: A Tool for Writing Cleaner, More Transparent Code

### 2020-02-04 joss/paper.Rmd
edits and spell check

At the core of the growth of scientific computation, the ``R``
statistical programming language has grown exponentially to become one
of the top ten programming languages in use today. At it's root R is a
*statistical* programming language. That is, it was designed for use
in analytical workflows; and the majority of the R community is
focused on producing code for idiosyncratic projects that are
*results* oriented. Also, R's design is intentionally at a level that
abstracts many aspects of programming that would otherwise act as a
barrier to entry for many users. This is good in that there are many
people who use R to their benefit with little to no formal training in
computer science or software engineering, but these same users can
also be frequently frustrated by code that is fragile, buggy and
complicated enough to quickly become obtuse even to the authors. The
stability, reproducibility and re-use of scientific analyses in R
would be improved by refactoring. From this perspective, tools that
can lower the time and energy required to re-factor and streamline
analytical scripts and otherwise help to "clean" code, but abstracted
enough to be easily accessible, could have a significant impact on
scientific reproducibility across all disciplines [@Visser2015].

@@

In the process of cleaning a script, it is likely that a user might
require some help analyzing the script. There are several functions to
help with this process. The `get_vars` function will return a list of
possible results for a given script at a supplied file path. This is
obviously an important step, and justifiably, the default behavior of
the `clean` function is to run `get_vars` if no results are
supplied. To help with limiting and checking the selection of results,
the `code_graph` function creates a network graph of the relationships
among the various results and lines of code in the script. Last, the
`get_libs` function can be used to detect the packages that a given
script depends on, which it will return as coded library calls that
can be inserted into a cleaned script.

@@

All of these processes rely on the generation of data provenance.  The
term provenance means information about the origins of some
object. Data provenance is a formal representation of the execution of
a computational process (https://www.w3.org/TR/prov-dm/), to
rigorously determine the the unique computational pathway from inputs
to results [@Carata2014]. To avoid confusion, note that "data" in this
context is used in a broad sense to include all of the information
generated during computation, not just the data that are collected in
a research project that are used as input to an analysis. Having the
formalized, mathematically rigorous representation that data
provenance provides guarantees that the analyses that *RClean*
conducts are theoretically sound. Most importantly, because the
relationships defined by the provenance can be represented as a graph,
it is possible to apply network search algorithms to determine the
minimum and sufficient code needed to generate the chosen result in
the `clean` function.

There are multiple approaches to collecting data provenance, but
`Rclean` uses "prospective" provenance, which analyzes code and uses
language specific information to predict the relationship among
processes and data objects. `Rclean` relies on a library called
`CodeDepends` to gather the prospective provenance for each
script. For more information on the mechanics of the `CodeDepends`
package, see [@Lang2019]. To get an idea of what data provenance is,
take a look at the `code_graph` function. The plot that it generates
is a graphical representation of the prospective provenance generated
for `Rclean`.

@@

```{R prov-graph, fig.cap = "Network diagram of the prospective data provenance generated for an example script. Arrows indicate relationships among objects and lines of code (numbered)."}

@@

Relatedly, it is important to briefly point out that `Rclean` *does
not* keep comments present in code. This could be seen as a limitation
of the data provenance, which currently does not assign them a
relationships. Therefore, although there is often very useful or even
invaluable information in comments, the `clean` function removes
them. This is a general issue with automated methods for detecting the
relationships between comments and the code itself. Comments at the
end of lines are typically relevant to the line they are on, but this
is not a requirement and could refer to other lines. Also, comments
occupying their own lines usually refer to the following lines, but
this is also not necessarily the case. As `clean` depends on the
unambiguous determination of relationships in the production of
results, it cannot operate automatically on comments. However,
comments in the original code remain untouched and can be used to
inform the reduced code. Also, as the `clean` function is oriented
toward isolating code based on a specific result, the resulting code
tends to naturally support the generation of new comments that are
higher level (e.g. "The following produces a plot of the mean response
of each treatment group."), and lower level comments are not necessary
because the code is simpler and clearer.

@@

Conducting analyses is challenging in that it requires thinking about
multiple concepts at the same time. What did I measure? What analyses
are relevant to them? Do I need to transform the data? What's the code
for the analysis I want to run? And so on. Data analysis can be messy
and complicated, so it's no wonder that the code often reflects
this. This is why having a way to isolate code based on variables can
be valuable. The following is an example of a script that has some
complications. As you can see, although the script is not extremely
long or complicated, it's difficult enough to make it frustrating to
visualize it in its entirety and pick through it.

@@

So, let's say we've come to our script wanting to extract the code to
produce one of the results `fit.sqrt.A`, which is an analysis that is
relevant to some product. Not only do we want to double check the
results, we also want to use the code again for another purpose, such
as creating a plot of the patterns supported by the test.

Manually tracing through our code for all the variables used in the
test and finding all of the lines that were used to prepare them for
the analysis would be annoying and difficult, especially given the
fact that we have used "x" as a prefix for multiple unrelated objects
in the script. Instead, we can easily do this automatically with
`Rclean`.

@@

We have created ``Rclean`` to provide a simple, easy to use tool for
scientists who would like help refactoring their code. Using
``Rclean`` the code necessary to produce a specified result (e.g., an
object stored in memory or a table or figure written to disk) can be
easily isolated even when tangled with code for other results. This
functionality is enabled by graph analysis of data provenance
collected from the user's script. As statistical programming becomes
more common across the sciences, tools that make the production of
accessible code will be an important aid for improving scientific
reproducibility.

@@

Also, future work could extend the existing framework to support new
and different provenance capture methods. One possibility is
*retrospective provenance*, which tracks a computational process as it
is executing. Through this active, concurrent monitoring,
retrospective provenance can gather information that static
prospective provenance can't. Only using prospective provenance means
that the outcomes of some processes can not be predicted. For example,
if there is a part of a script that is determined by a random number,
the current implementation of prospective provenance can not predict
the path that will be taken through the code. Therefore, the code
cannot be reduced to exclude the pathway that would not be
taken. However, using retrospective provenance comes at a cost. In
order to gather it, the script needs to be executed. When scripts are
computationally intensive or contain bugs that stop execution, then
retrospective provenance can not be obtained for part or all of the
code. Some work has already been done in the direction of implementing
retrospective provenance for code cleaning in R (see
http://end-to-end-provenance.github.io).

>>>>

At the core of the growth of scientific computation, the ``R``
statistical programming language has grown exponentially to become one
of the top ten programming languages in use today. At it's root R is a
*statistical* programming language. That is, it was designed for use
in analytical workflows; and the majority of the R community is
focused on producing code for idiosyncratic projects that are
*results* oriented. Also, R's design is intentionally at a level that
abstracts many aspects of programming that would otherwise act as a
barrier to entry for many users. This is good in that there are many
people who use R to their benefit with little to no formal training in
computer science or software engineering, but these same users can
also be frequently frustrated by code that is fragile, buggy and
complicated enough to quickly become obtuse even to the authors. The
stability, reproducibility and re-use of scientific analyses in R
would be improved by refactoring, which is a common practice in
software engineering [@Martin2009]. From this perspective, tools that
can lower the time and energy required to refactor analytical scripts
and otherwise help to "clean" code, but abstracted enough to be easily
accessible, could have a significant impact on scientific
reproducibility across all disciplines [@Visser2015].

@@

In the process of cleaning a script, it is likely that a user will not
know the precise objects they want and might require some help
analyzing it. There are several functions to help with this
process. The `get_vars` function will return a list of possible
results for a given script at a supplied file path. This is obviously
an important step, and justifiably, the default behavior of the
`clean` function is to run `get_vars` if no results are supplied. To
help with limiting and checking the selection of results, the
`code_graph` function creates a network graph of the relationships
among the various results and lines of code in the script, which is
the function used to create the figure demonstrating data provenance
\ref{prov-graph}. Last, the `get_libs` function can be used to detect
the packages that a given script depends on, which it will return as
coded library calls that can be inserted into a cleaned script.

@@

All of these processes rely on the generation of data provenance.  The
term provenance means information about the origins of some
object. Data provenance is a formal representation of the execution of
a computational process (https://www.w3.org/TR/prov-dm/), to
rigorously determine the the unique computational pathway from inputs
to results [@Carata2014]. To avoid confusion, note that "data" in this
context is used in a broad sense to include all of the information
generated during computation, not just the data that are collected in
a research project that are used as input to an analysis. Having the
formalized, mathematically rigorous representation that data
provenance provides guarantees that the analyses that `Rclean`
conducts are theoretically sound. Most importantly, because the
relationships defined by the provenance can be represented as a graph,
it is possible to apply network search algorithms to determine the
minimum and sufficient code needed to generate the chosen result in
the `clean` function.

There are multiple approaches to collecting data provenance, but
`Rclean` uses "prospective" provenance, which analyzes code and uses
language specific information to predict the relationship among
processes and data objects. `Rclean` relies on a library called
`CodeDepends` to gather the prospective provenance for each
script. For more information on the mechanics of the `CodeDepends`
package, see [@Lang2019]. To get an idea of what data provenance is,
take a look at the `code_graph` function. The plot that it generates
is a graphical representation of the prospective provenance generated
for `Rclean` \ref{prov-graph}.

@@

```{R prov-graph, fig.cap = "Network diagram of the prospective data
provenance generated for an example script. Arrows indicate
which lines of code (numbered) produced which objects (named).", echo = FALSE}

@@

Relatedly, it is important to point out that `Rclean` *does not* keep
comments present in code. This could be seen as a limitation of the
data provenance, which currently does not assign them a
relationships. Therefore, although there is often very useful or even
invaluable information in comments, the `clean` function removes
them. This is a general issue with automated methods for detecting the
relationships between comments and the code itself. Comments at the
end of lines are typically relevant to the line they are on, but this
is not a requirement and could refer to other lines. Also, comments
occupying their own lines usually refer to the following lines, but
this is also not necessarily the case. As `clean` depends on the
unambiguous determination of relationships in the production of
results, it cannot operate automatically on comments. However,
comments in the original code remain untouched and can be used to
inform the reduced code. Also, as the `clean` function is oriented
toward isolating code based on a specific result, the resulting code
tends to naturally support the generation of new comments that are
higher level (e.g. "The following produces a plot of the mean response
of each treatment group."), and lower level comments are not necessary
because the code is simpler and clearer.

@@

Data analysis can be messy and complicated, so it's no wonder that the
code often reflects this. "What did I measure? What analyses are
relevant to them? Do I need to transform the data? What's the function
for the analysis I want to run?" This is why having a way to isolate
code based on variables can be valuable. The following is an example
of a script that has some complications. As you can see, although the
script is not extremely long or complicated, it's difficult enough to
make it frustrating to visualize it in its entirety and pick through
it.

@@

So, let's say we've come to our script wanting to extract the code to
produce one of the results `fit.sqrt.A`, which is an analysis that is
relevant to some product. Not only do we want to double check the
results, we also want to use the code again for another purpose, such
as creating a plot of the patterns supported by the test. Manually
tracing through our code for all the variables used in the test and
finding all of the lines that were used to prepare them for the
analysis would be annoying and difficult, especially given the fact
that we have used "x" as a prefix for multiple unrelated objects in
the script. Instead, we can easily do this automatically with
`Rclean`.

@@

We have created ``Rclean`` to provide a simple, easy to use tool for
scientists who would like help refactoring code. Using ``Rclean`` the
code necessary to produce a specified result (e.g., an object stored
in memory or a table or figure written to disk) can be easily and
*reliably* isolated even when tangled with code for other
results. This functionality is enabled by graph analysis of data
provenance collected from the user's script. This is likely to be a
broadly useful tool as statistical programming becomes more common
across the sciences. Tools, such as this, that make the production of
accessible code will be an important aid for improving scientific
reproducibility.

@@

Another possible future direction could extend the existing framework
to support new and different provenance capture methods. One
possibility is *retrospective provenance*, which tracks a
computational process as it is executing. Through this active,
concurrent monitoring, retrospective provenance can gather information
that static prospective provenance can't. Only using prospective
provenance means that the outcomes of some processes can not be
predicted. For example, if there is a part of a script that is
determined by a random number, the current implementation of
prospective provenance can not predict the path that will be taken
through the code. Therefore, the code cannot be reduced to exclude the
pathway that would not be taken. However, using retrospective
provenance comes at a cost. In order to gather it, the script needs to
be executed. When scripts are computationally intensive or contain
bugs that stop execution, then retrospective provenance can not be
obtained for part or all of the code. Some work has already been done
in the direction of implementing retrospective provenance for code
cleaning in R (see http://end-to-end-provenance.github.io).
